---
title: "Differential expression"
output: html_document
---

```{r, include=FALSE}
source('functions.R')
library(org.Hs.eg.db)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(limma)
```


```{r , echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, include=FALSE}

df_mRNA <- read.table("feature_genes.tsv", sep = "\t", header = TRUE, row.names = 1)
colnames(df_mRNA) <- gsub(".", "-", x = colnames(df_mRNA), fixed = T)

meta_data <- read.table("datasheet.csv", sep=",", header = TRUE)

rownames(meta_data) <- meta_data$name
df_mRNA = df_mRNA[,rownames(meta_data)]

all(rownames(meta_data) %in% colnames(df_mRNA))

data <- df_mRNA %>%  rownames_to_column()
```


```{r , include = FALSE}
# set up meta_data for all experiments

meta_data_1 <- meta_data %>% 
  filter(Time == "24hr") %>% 
  column_to_rownames("name")

meta_data_2 <- meta_data %>% 
  filter(Time == "3hr") %>% 
  column_to_rownames("name")

meta_data_3 <- meta_data %>% 
  filter(Time == "6hr") %>% 
  filter(Sample != "2" | Treatment != "DMSO") %>% 
  column_to_rownames("name")


```




```{r , message=FALSE, include=FALSE}
## Set up the DESeq2 object
res_24 <- run_deseq2(as.data.frame(df_mRNA), meta_data_1, control = "DMSO", test="J4", value="Treatment")
res_3 <- run_deseq2(as.data.frame(df_mRNA), meta_data_2, control = "DMSO", test="J4", value="Treatment")
res_6 <- run_deseq2(as.data.frame(df_mRNA), meta_data_3, control = "DMSO", test="J4", value="Treatment")
```

The following describes the analysis of the data using DEseq2. I have seperated the analysis into the four different contrasts and have plotted summary statistics and MA plots for each.  


## Summary of the data {.tabset .tabset-fade}

### 24hr J4 vs DMSO

```{r , echo=FALSE}
summary(res_24, alpha=0.05)
```

### 6hr J4 vs DMSO

```{r , echo=FALSE}
summary(res_6, alpha=0.05)
```

### 3hr J4 vs DMSO

```{r , echo=FALSE}
summary(res_3, alpha=0.05)
```

## MA plots {.tabset .tabset-fade}

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored red if the adjusted p value is less than 0.01. Points which fall out of the window are plotted as open triangles pointing either up or down.


### 24 J4 vs DMSO

```{r }
DESeq2::plotMA(res_24)
```

### 6hr J4 vs DMSO

```{r }
DESeq2::plotMA(res_6)
```

### 3hr J4 vs DMSO

```{r }
DESeq2::plotMA(res_3)
```

# Volcano plots {.tabset .tabset-fade}

## 24hr J4 vs DMSO

```{r}

library("ggplot2") #Best plots
library("ggrepel") #Avoid overlapping labels


test <- as.data.frame(res_24)
  
data <- as.vector(rownames(test))
annots <-  AnnotationDbi::select(org.Hs.eg.db, keys=data,
                                   columns="SYMBOL", keytype = "ENSEMBL")
  
result <- merge(test, annots, by.x="row.names", by.y="ENSEMBL")
res <- result %>% 
    dplyr::select(log2FoldChange, SYMBOL, baseMean, padj, Row.names) %>% 
    na.omit()
  

mutateddf <- mutate(res, sig=ifelse(res$padj<0.01, "padj<0.01", "Not Sig")) #Will have different colors depending on significance
input <- cbind(gene=rownames(mutateddf), mutateddf )
input <- input %>% 
  arrange(input$padj)

symbol_data <- head(input, 30)

#convert the rownames to a column
volc = ggplot(input, aes(log2FoldChange, -log10(padj))) + #volcanoplot with log2Foldchange versus pvalue
    geom_point(aes(col=sig)) + #add points colored by significance
geom_point(data=symbol_data, aes(log2FoldChange, -log10(padj)), colour="red") +
      ggtitle("MUG KDOBA67") #e.g. 'Volcanoplot DESeq2'

#setEPS()
#postscript("MUG_volcano.eps")
volc+geom_text_repel(data=symbol_data, aes(label=`SYMBOL`)) + scale_colour_Publication() + theme_bw()#adding text for the genes
#ggsave("Volcanoplot.jpeg", device="jpeg") #In case you want to easily save to disk
#dev.off()

```


## 6hr J4 vs DMSO

```{r}

library("ggplot2") #Best plots
library("ggrepel") #Avoid overlapping labels


test <- as.data.frame(res_6)
  
data <- as.vector(rownames(test))
annots <-  AnnotationDbi::select(org.Hs.eg.db, keys=data,
                                   columns="SYMBOL", keytype = "ENSEMBL")
  
result <- merge(test, annots, by.x="row.names", by.y="ENSEMBL")
res <- result %>% 
    dplyr::select(log2FoldChange, SYMBOL, baseMean, padj, Row.names) %>% 
    na.omit()
  

mutateddf <- mutate(res, sig=ifelse(res$padj<0.01, "padj<0.01", "Not Sig")) 


#Will have different colors depending on significance
input <- cbind(gene=rownames(mutateddf), mutateddf )
input <- input %>% 
  arrange(input$padj)

symbol_data <- head(input, 10)


#convert the rownames to a column
volc = ggplot(input, aes(log2FoldChange, -log10(padj))) + #volcanoplot with log2Foldchange versus pvalue
    geom_point(aes(col=sig)) + 
    geom_point(data=symbol_data, aes(log2FoldChange, -log10(padj)), colour="red") + #add points colored by significance
    ggtitle("UCH1 KDOBA67") #e.g. 'Volcanoplot DESeq2'

    



#setEPS()
#postscript("UCH1_volcano_UCH7.eps")
volc+geom_text_repel(data=symbol_data, aes(label=`SYMBOL`)) + scale_colour_Publication() + theme_bw()#adding text for the genes
#ggsave("Volcanoplot.jpeg", device="jpeg") #In case you want to easily save to disk
#dev.off()
#volc

```

## 3hr J4 vs DMSO

```{r}

library("ggplot2") #Best plots
library("ggrepel") #Avoid overlapping labels


test <- as.data.frame(res_3)
  
data <- as.vector(rownames(test))
annots <-  AnnotationDbi::select(org.Hs.eg.db, keys=data,
                                   columns="SYMBOL", keytype = "ENSEMBL")
  
result <- merge(test, annots, by.x="row.names", by.y="ENSEMBL")
res <- result %>% 
    dplyr::select(log2FoldChange, SYMBOL, baseMean, padj, Row.names) %>% 
    na.omit()
  

mutateddf <- mutate(res, sig=ifelse(res$padj<0.01, "padj<0.01", "Not Sig")) #Will have different colors depending on significance
input <- cbind(gene=rownames(mutateddf), mutateddf )
input <- input %>% 
  arrange(input$padj)

symbol_data <- head(input, 10)

#convert the rownames to a column
volc = ggplot(input, aes(log2FoldChange, -log10(padj))) + #volcanoplot with log2Foldchange versus pvalue
    geom_point(aes(col=sig)) + #add points colored by significance
geom_point(data=symbol_data, aes(log2FoldChange, -log10(padj)), colour="red") +
      ggtitle("UM KDOBA67") #e.g. 'Volcanoplot DESeq2'

#setEPS()
#postscript("UM_volcano.eps")
volc+geom_text_repel(data=symbol_data, aes(label=`SYMBOL`)) + scale_colour_Publication() + theme_bw()#adding text for the genes
#ggsave("Volcanoplot.jpeg", device="jpeg") #In case you want to easily save to disk
#dev.off()

```

## Results tables

The folowing results tables show the significant genes. Filtering has been performed with a log2 fold change +/- 2.


### 24hr J4 vs DMSO

```{r , message=FALSE, echo=FALSE}
dt <- filter_genes(as.data.frame(res_24), name="featurecounts_24hrJ4vsdmso")

datatable(dt$sig)
```

### 6hr J4 vs DMSO

```{r , message=FALSE, echo=FALSE}
dt <- filter_genes(as.data.frame(res_6), name="featurecounts_6hrJ4vsdmso")

datatable(dt$sig)
```

### 3hr J4 vs DMSO

```{r , message=FALSE, echo=FALSE}
dt <- filter_genes(as.data.frame(res_3), name="featurecounts_3hrJ4vsdmso")

datatable(dt$sig)
```


# Conclusion

Given the high level of low count genes in the dataset: The biological and technical variation in the experiment dominates any true log fold changes across condition given your sample size. Another way to say this is that the experiment was underpowered to detect the changes across condition.